A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-15 latex/main.tex
Tweaks

Millions of SARS-CoV-2 genome sequences were collected in
response to the COVID-19 pandemic, forming a dataset of unprecedented
richness.
Estimated genealogies are fundamental to
gleaning information from this ocean of data, 
and form the primary input to many downstream analyses.
A basic assumption of methods to infer genealogies from viral 
genetic sequence data is that recombination is negligible and 
the genealogy is a tree.
Recombinant lineages have risen to global prevalence in SARS-CoV-2,
and therefore simple tree representations are necessarily
incomplete and potentially misleading.
We present sc2ts, a method to infer reticulate genealogies 
as an Ancestral Recombination Graph (ARG) for SARS-CoV-2
that automatically detects and incorporates recombination.  
We use sc2ts to infer an ARG of 2.48 million genomes,
and show that it captures a broad range of 
known evolutionary features and provides novel 
insights into the relationships between previously identified
recombinant lineages.
Together, the ARG, inference method and 
suite of open-source analysis tools, 
represent a powerful resource for future research.

@@

The global
surveillance data collected during the COVID-19 pandemic presents fundamental challenges to existing approaches
in two ways. Firstly, the sheer amount of data 
overwhelmed even approximate distance-matrix based methods.
% Accessed GISAID/EpiCov on 08 Oct., 2025
At the peak in February, 2022, about 80 thousand samples
were submitted on average per day to GISAID\cite{Shu2017},
accumulating to a total of over 17.5 million samples
as of October, 2025.
Regularly reinferring trees de-novo from such rapidly
growing datasets is untenable.
The UShER\cite{Turakhia2021} method is based on adding
samples to an existing tree using maximum parsimony,
with periodic rearrangements to counter the inherent 
greediness of the approach\cite{DeMaio2023}.
The benefits of this simple model and the value of
a comprehensive genealogy updated in real-time~\cite{McBroome2021}
were conclusively demonstrated, as the resource
quickly became an indispensable element
of the pandemic response\cite{Hinrichs2024}.
The second challenge posed is that recombination is an important force
in SARS-CoV-2 evolution \cite{Jackson2021,VanInsberghe2021,Focosi2022,
Gutierrez2022,Ignatieva2022,Tamura2023},
with multiple deep recombination events now known to be present
in the genealogy of the global population \cite{Roemer2023,Shiraz2023enhanced}.
Although several methods have been proposed to \emph{detect}
recombinant samples~\cite{
Varabyou2021rapid,% Bolotie
Turakhia2022,%RIPPLES
Zhou2023virusrecom,% CovRecomb
Alfonsi2024,%RecombinHunt 
Li2024CovRecomb%CovRecomb 
},
this information is not incorporated into the phylogenies used 
in innumerable studies~\cite{McBroome2021,Hadfield2018},
and recombination continues to be treated as a 
nuisance factor.

@@

Here we present sc2ts, a new method to infer ARGs for SARS-CoV-2  
that incrementally adds samples over time in a similar manner 
to UShER, but also automatically detects and incorporates recombination
in real-time. Based on the Viridian dataset, a freely available  
resource purged of various systematic errors and artifacts \cite{Hunt2024},
we infer an ARG with 2.48 million whole genome sequences.
We show that this ARG is highly congruent with the UShER phylogeny built 
from the same dataset as well as the Pango nomenclature system\cite{Rambaut2020},
and also accurately captures evolutionary features such as mutational 
spectra\cite{Bloom2023} and deletions\cite{Li2023indels}.
We show that sc2ts automatically and accurately detects details about
known recombinant lineages that were painstakingly collated in a community 
process, and provides novel insights into the relationships between these
recombinant lineages. We compare the recombination events detected by 
sc2ts with the state-of-the-art RIPPLES method~\cite{Turakhia2022}, showing
that sc2ts is more sensitive and has a substantially lower false positive rate.
The output of sc2ts---incorporating
genealogical relationships, point mutations, deletions, and 
recombination---provides a basis for the study of SARS-CoV-2
that systematically accounts for all of these evolutionary forces.
This ARG, coupled with the rich suite of supporting software
in the tskit\cite{Ralph2020} and VCF-Zarr~\cite{Czech2025} ecosystems,
enables pandemic-scale analysis on a laptop. 
These resources
will help us to learn from the vast troves of data collected during the
pandemic, and to prepare for the next.

@@

Sc2ts reconstructs an ARG by incrementally adding daily batches of 
sequences (Figure~\ref{fig:methodology}).
First, sc2ts infers likely ``copying paths'' connecting each sample in the 
batch to node(s) in the current ARG
using a simplified version of the Li and Stephens (LS) model\cite{LiStephens2003}. 
% Important to go into a bit of detail about LS here as it's so fundamental
% and novel in the area.
The LS model is a Hidden Markov Model (HMM) widely used in human genomics
to approximate the effects of mutation and recombination on genetic
inheritance\citep{Delaneau2019-wl,Browning2018-nk},
typically parameterised by population-scaled mutation and
recombination rates varying along the genome~\cite{Donnelly2010-coalescent}.
In the interest of interpretability, we simplified this model to 
use a single parameter $k$ which essentially 
controls the number of recurrent mutations that is preferred to 
switching to a different parent via recombination (we use $k=4$ here).
We used the highly efficient LS HMM implementation in tsinfer\cite{Kelleher2019},
with some additional optimisations enabled by our simplified model,
to find exact Viterbi solutions of these copying paths for each sample
given the current ARG.
Note that this is a form of parsimony, with the recombination
penalty $k$ defining the relative parsimoniousness of recombinant
vs non-recombinant solutions to the HMM.
The per-sample copying paths produced by the HMM (most of which do not 
involve recombination) form clusters of samples within a batch, which 
are then resolved using standard phylogenetic techniques.
Finally, sc2ts attaches the phylogenies of the clusters to the current ARG, and
applies parsimony-improving heuristics to address issues introduced 
by the inherent greediness of this strategy.
Sc2ts is implemented in Python,
building on core PyData infrastructure~\cite{Harris2020array,Virtanen2020scipy,
Mckinney2010data,Lam2015numba} and bioinformatics 
packages\cite{Shirley2015efficient,Kunzmann2018biotite,Kunzmann2023biotite}.
See STAR Methods for details on all aspects of the method.

@@

% NOTE: Trying to clarify that this of the pandemic period and we're 
% not trying to cover *all* history.
The ARG is a comprehensive and deeply validated record of the pandemic phase 
of SARS-CoV-2 evolutionary history, accurately reflecting
a range of results from the literature.
First, we compared the sc2ts ARG and with the state-of-the-art UShER 
tree inferred from the same dataset
by pruning both to their intersection of 2,475,418 samples and 27,507 sites
(STAR Methods)
% JK: this feels a bit wishy washy, but there really is no simple answer to
% this and I don't want to claim we're more parsimonious.
and found they were very similar in terms of parsimony
(Document~\ref{sec:supplement_sc2ts_usher_parsimony})
and imputation of missing and ambiguous bases
(Document~\ref{sec:supplement_sc2ts_usher_imputation}).
We also compared the mutational spectra of major VOCs in the ARG
against the UShER tree and 
a previous study \cite{Bloom2023} (STAR Methods),
finding close agreement (Figure~\ref{fig:mutational_spectra}).
We validated that the broad phylogenetic structure of the ARG and 
UShER tree closely match by generating tanglegrams for a set of representative
samples from 814 deeply-sampled lineages (STAR Methods; 
Figures~\ref{fig:tanglegram_base},
\ref{fig:tanglegram_delta},
\ref{fig:tanglegram_ba2},
\ref{fig:tanglegram_ba5}).
We also evaluated how well the ARG reflects the structure
of the Pango nomenclature system by computing
Pango assignments for all 2.75 million nodes in the ARG using
Pangolin~\cite{OToole2021assignment} (STAR Methods).
All but 21,680 of the exported sample alignments (0.87\%) agreed
with the source consensus sequences in terms of Pango lineage assignment
(comparable to the 0.27\% disagreement between pangolin-data
versions 1.21 and 1.29 in the source Viridian metadata).
A majority of these lineages (1473 of 2058)
have monophyletic origins within the ARG,
increasing to 1779 (86\%) when we allow
for multiple sibling origination nodes
(Document~\ref{sec:supplement_pango_lineages}).
The ARG is time-resolved, with internal node dates estimated by
a non-parametric method~\cite{Pope2025} (STAR Methods). 
We show that the dates agree reasonably with NextStrain 
estimates (Figure \ref{fig:pango_vs_nextstrain_node_times})
and overall accuracy to another large-scale
method~\cite{sanderson2021chronumental} (Document \ref{sec:evaluating_node_dates}).
Finally, we verified
that the origins of the 
Alpha (Figure~\ref{fig:subgraph_alpha}, Document~\ref{sec:dd_alpha}),
Delta (Figure~\ref{fig:subgraph_delta}, Document~\ref{sec:dd_delta})
and
Omicron (Figure~\ref{fig:subgraph_omicron}, Document~\ref{sec:dd_omicron})
VOCs reflect expectations from the literature\cite{RambautAlpha2020,ECDCreport,
Cherian2021,Farinholt2021,Stern2021,Dor2025,RambautOmicron2021}.

Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023},
but are not incorporated into UShER trees and can be difficult
to account for. Because deletions are difficult to distinguish 
from bioinformatics errors
and their multi-base nature presents challenges
to the Markovian structure of the LS HMM,
we masked out all gap characters as missing data during primary inference.
To approximate the effects of the most important deletions into the ARG,
we remapped data using post hoc parsimony at 75 sites involved in deletions
with a frequency of $\geq 1\%$ reported by Li et al.\cite{Li2023indels}
during post-processing (STAR Methods).
Although sites are treated independently in this approach,
runs of mutations to the gap character at adjacent sites
correctly identified many known
high-frequency deletions (Table~\ref{tab:major_dels}) and
recurrent deletions (Table~\ref{tab:rec_dels}).
For example, the three characteristic deletions of Alpha \cite{RambautAlpha2020}
(11288+9, 21765+6, 21991+3,
where the first and second numbers are
the start position and length, respectively)
are correctly mapped above the Alpha origin in the ARG
(Figure~\ref{fig:subgraph_alpha}).
The ARG captures known recurrent deletions
in the N-terminal domain of the S1 unit of Spike
(which is a deletion hotspot \cite{McCarthy2021})
for example,
the 21765+6 deletion ($\Delta$H69/V70),
which causes S-Gene Target Failure \cite{Walker2021} and
is associated with increased efficiency in cell entry \cite{Meng2021}, and
the 21991+3 deletion
which is associated with antibody escape \cite{McCarthy2021}.
See Document~\ref{sec:supplement_dels} for further analysis.

>>>>

Millions of SARS-CoV-2 genome sequences were collected in
response to the COVID-19 pandemic, forming a dataset of unprecedented
richness.
Estimated genealogies are fundamental to
gleaning information from this ocean of data, 
and form the primary input to many downstream analyses.
A basic assumption of methods to infer genealogies from viral 
genetic sequence data is that recombination is negligible and 
the genealogy is a tree.
Recombinant lineages have risen to global prevalence in SARS-CoV-2,
and simple tree representations are therefore necessarily
incomplete and potentially misleading.
We present sc2ts, a method to infer reticulate genealogies 
as an Ancestral Recombination Graph (ARG) for SARS-CoV-2
that automatically detects and incorporates recombination.  
We use sc2ts to infer an ARG of 2.48 million genomes,
and show that it captures a broad range of 
known evolutionary features and provides novel 
insights into the relationships between previously identified
recombinant lineages. [NOTE: mention overall estimate
of recombination compared with UShER+RIPPLES?]
Together, the ARG, inference method and 
suite of open-source analysis tools, 
represent a powerful resource for future research.

@@

The global
surveillance data collected during the COVID-19 pandemic presents fundamental challenges to existing approaches
in two ways. Firstly, the sheer amount of data 
overwhelmed even approximate distance-matrix based methods.
% Accessed GISAID/EpiCov on 08 Oct., 2025
At the peak in February 2022 about 80 thousand samples
per day  were submitted to GISAID\cite{Shu2017},
accumulating to a total of over 17.5 million samples
as of October 2025.
Regularly reinferring trees de-novo from such rapidly
growing datasets is untenable.
The UShER\cite{Turakhia2021} method is based on adding
samples to an existing tree using maximum parsimony,
with periodic rearrangements to counter the inherent 
greediness of the approach\cite{DeMaio2023}.
The benefits of this simple model and the value of
a comprehensive genealogy updated in real-time~\cite{McBroome2021}
were conclusively demonstrated, as the resource
quickly became an indispensable element
of the pandemic response\cite{Hinrichs2024}.
The second challenge posed is that recombination is an important force
in SARS-CoV-2 evolution \cite{Jackson2021,VanInsberghe2021,Focosi2022,
Gutierrez2022,Ignatieva2022,Tamura2023},
with multiple deep recombination events now known to be present
in the genealogy of the global population \cite{Roemer2023,Shiraz2023enhanced}.
Although several methods have been proposed to \emph{detect}
recombinant samples~\cite{
Varabyou2021rapid,% Bolotie
Turakhia2022,%RIPPLES
Zhou2023virusrecom,% CovRecomb
Alfonsi2024,%RecombinHunt 
Li2024CovRecomb%CovRecomb 
},
this information is not incorporated into the phylogenies used 
in innumerable studies~\cite{McBroome2021,Hadfield2018},
and recombination continues to be treated as a 
nuisance factor.

@@

Here we present sc2ts, a new method to infer ARGs for SARS-CoV-2  
that incrementally adds samples over time in a similar manner 
to UShER, but also automatically detects and incorporates recombination
in real-time. Based on the Viridian dataset, a freely available  
resource purged of various systematic errors and artifacts \cite{Hunt2024},
we infer an ARG with 2.48 million whole genome sequences.
We show that this ARG is highly congruent with the UShER phylogeny built 
from the same dataset as well as the Pango nomenclature system\cite{Rambaut2020},
and also accurately captures evolutionary features such as mutational 
spectra\cite{Bloom2023} and deletions\cite{Li2023indels}.
We show that sc2ts automatically and accurately detects details about
known recombinant lineages that were painstakingly collated in a community 
process, and provides novel insights into the relationships between these
recombinant lineages. We compare the recombination events detected by 
sc2ts with the state-of-the-art combination of UShER and 
RIPPLES ~\cite{Turakhia2022}, showing that sc2ts is more sensitive,
detecting more of the well-characterized recombination events 
curated by the Pango lineage designation community.
The output of sc2ts---incorporating
genealogical relationships, point mutations, deletions, and 
recombination---provides a basis for the study of SARS-CoV-2
that systematically accounts for all of these evolutionary forces.
This ARG, coupled with the rich suite of supporting software
in the tskit\cite{Ralph2020} and VCF-Zarr~\cite{Czech2025} ecosystems,
enables pandemic-scale analysis on a laptop. 
These resources
will help us to learn from the vast troves of data collected during the
pandemic, and to prepare for the next.

@@

Sc2ts reconstructs an ARG by incrementally adding daily batches of 
sequences (Figure~\ref{fig:methodology}).
First, sc2ts infers likely ``copying paths'' connecting each sample in the 
batch to node(s) in the current ARG
using a simplified version of the Li and Stephens (LS) model\cite{LiStephens2003}. 
% Important to go into a bit of detail about LS here as it's so fundamental
% and novel in the area.
The LS model is a Hidden Markov Model (HMM) widely used in human genomics
to approximate the effects of mutation and recombination on genetic
inheritance\citep{Delaneau2019-wl,Browning2018-nk},
typically parameterised by population-scaled mutation and
recombination rates varying along the genome~\cite{Donnelly2010-coalescent}.
In the interest of interpretability, we simplified this model to 
use a single parameter $k$ which essentially 
controls the number of recurrent mutations that is preferred to 
switching to a different parent via recombination (we use $k=4$ here
[TODO cite support for this, like CovRecomb etc]).
We used the highly efficient LS HMM implementation in tsinfer\cite{Kelleher2019},
with some additional optimisations enabled by our simplified model,
to find exact Viterbi solutions of these copying paths for each sample
given the current ARG.
Note that this is a form of parsimony, with the recombination
penalty $k$ defining the relative parsimoniousness of recombinant
vs non-recombinant solutions to the HMM.
The per-sample copying paths produced by the HMM (most of which do not 
involve recombination) form clusters of samples within a batch, which 
are then resolved using standard phylogenetic techniques.
Finally, sc2ts attaches the phylogenies of the clusters to the current ARG, and
applies parsimony-improving heuristics to address issues introduced 
by the inherent greediness of this strategy.
Sc2ts is implemented in Python,
building on core PyData infrastructure~\cite{Harris2020array,Virtanen2020scipy,
Mckinney2010data,Lam2015numba} and bioinformatics 
packages\cite{Shirley2015efficient,Kunzmann2018biotite,Kunzmann2023biotite}.
See STAR Methods for details on all aspects of the method.

@@

% NOTE: Trying to clarify that this of the pandemic period and we're 
% not trying to cover *all* history.
The ARG is a comprehensive and deeply validated record of the pandemic phase 
of SARS-CoV-2 evolutionary history, accurately reflecting
a range of results from the literature.
First, we compared the sc2ts ARG and with the state-of-the-art UShER 
tree inferred from the same dataset
by pruning both to their intersection of 2,475,418 samples and 27,507 sites
(STAR Methods)
% JK: this feels a bit wishy washy, but there really is no simple answer to
% this and I don't want to claim we're more parsimonious.
and found they were very similar in terms of parsimony
(Document~\ref{sec:supplement_sc2ts_usher_parsimony})
and imputation of missing and ambiguous bases
(Document~\ref{sec:supplement_sc2ts_usher_imputation}).
We also compared the mutational spectra of major VOCs in the ARG
against the UShER tree and 
a previous study \cite{Bloom2023} (STAR Methods),
finding close agreement (Figure~\ref{fig:mutational_spectra}).
We validated that the broad phylogenetic structure of the ARG and 
UShER tree closely match by generating tanglegrams for a set of representative
samples from 814 well-sampled lineages (STAR Methods; 
Figures~\ref{fig:tanglegram_base},
\ref{fig:tanglegram_delta},
\ref{fig:tanglegram_ba2},
\ref{fig:tanglegram_ba5}).
We also evaluated how well the ARG reflects the structure
of the Pango nomenclature system by computing
Pango assignments for all 2.75 million nodes in the ARG using
Pangolin~\cite{OToole2021assignment} (STAR Methods).
All but 21,680 of the exported sample alignments (0.87\%) agreed
with the source consensus sequences in terms of Pango lineage assignment
(comparable to the 0.27\% disagreement between pangolin-data
versions 1.21 and 1.29 in the source Viridian metadata).
A majority of these lineages (1473 of 2058)
have monophyletic origins within the ARG,
increasing to 1779 (86\%) when we allow
for multiple sibling origination nodes
(Document~\ref{sec:supplement_pango_lineages}).
The ARG is time-resolved, with internal node dates estimated by
a non-parametric method~\cite{Pope2025} (STAR Methods). 
We show that the dates agree reasonably with NextStrain 
estimates (Figure \ref{fig:pango_vs_nextstrain_node_times})
and overall accuracy comparable to 
Chronumental~\cite{sanderson2021chronumental} applied to 
an UShER tree (Document \ref{sec:evaluating_node_dates}).
Finally, we verified
that the origins of the 
Alpha (Figure~\ref{fig:subgraph_alpha}, Document~\ref{sec:dd_alpha}),
Delta (Figure~\ref{fig:subgraph_delta}, Document~\ref{sec:dd_delta})
and
Omicron (Figure~\ref{fig:subgraph_omicron}, Document~\ref{sec:dd_omicron})
VOCs reflect expectations from the literature\cite{RambautAlpha2020,ECDCreport,
Cherian2021,Farinholt2021,Stern2021,Dor2025,RambautOmicron2021}.

Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023},
but are not incorporated into UShER trees and must therefore
be accounted for separately.
Because deletions are difficult to distinguish 
from bioinformatics errors
and their multi-base nature presents challenges
to the Markovian structure of the LS HMM,
we masked out all gap characters as missing data during primary inference.
To approximate the effects of the most important deletions into the ARG,
we remapped data using post hoc parsimony at 75 sites involved in deletions
with a frequency of $\geq 1\%$ reported by Li et al.\cite{Li2023indels}
during post-processing (STAR Methods).
Although sites are treated independently in this approach,
runs of mutations to the gap character at adjacent sites
correctly identified many known
high-frequency deletions (Table~\ref{tab:major_dels}) and
recurrent deletions (Table~\ref{tab:rec_dels}).
For example, the three characteristic deletions of Alpha \cite{RambautAlpha2020}
(11288+9, 21765+6, 21991+3,
where the first and second numbers are
the start position and length, respectively)
are correctly mapped above the Alpha origin in the ARG
(Figure~\ref{fig:subgraph_alpha}).
The ARG captures known recurrent deletions
in the N-terminal domain of the S1 unit of Spike
(which is a deletion hotspot \cite{McCarthy2021})
for example,
the 21765+6 deletion ($\Delta$H69/V70),
which causes S-Gene Target Failure \cite{Walker2021} and
is associated with increased efficiency in cell entry \cite{Meng2021}, and
the 21991+3 deletion
which is associated with antibody escape \cite{McCarthy2021}.
See Document~\ref{sec:supplement_dels} for further analysis.
